---
title: "Student"
author: "Asela Gunatilake, 500534267"
date: "10/04/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggpubr)

```

# Loading data
2 datasets, 1 math, 1 portugese 

```{r}
d.mat=read.table("data/student/student-mat.csv",sep=";",header=TRUE)
d.port=read.table("data/student/student-por.csv",sep=";",header=TRUE)

d.all=merge(d.mat,d.port,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(nrow(d.all)) # 382 students

```

```{r}
head(d.mat)

```
Predictor levels
```{r}
unique(d.mat$school) #binary, two schools
unique(d.mat$age) #int, from 18 to 22 but most less than 20
hist(d.mat$age)
unique(d.mat$address) #binary, R:rural, U:urban
unique(d.mat$famsize) #binary, Greater than or equal to 3: GT3
unique(d.mat$Pstatus) #binany, parent status, living together - T
unique(d.mat$Medu) #Ordinal, o -4 scale, mother's education
unique(d.mat$Fedu)
unique(d.mat$Mjob) #nominal, 5 levels
unique(d.mat$Fjob)
unique(d.mat$reason) #nominal, 5 levels reason to select school
unique(d.mat$guardian) #nominal, 3 levels 
unique(d.mat$traveltime) #interval, 4 levels
unique(d.mat$studytime) #interval, 4 levels
unique(d.mat$failures) #ratio, max 4
unique(d.mat$schoolsup) #binary, extra school support 
unique(d.mat$famsup) #binary
unique(d.mat$paid) #binary, extra school support
unique(d.mat$activities) #binary, extra curricular
unique(d.mat$nursery) #binary, attended
unique(d.mat$higher) #binary, wants to attend higher edu
unique(d.mat$internet) #binary, int. access from home
unique(d.mat$romantic) #binary
unique(d.mat$famrel) #ordinal, quality of family relationship 1-5 levels
unique(d.mat$freetime) #ordinal, 1-5 levels
unique(d.mat$goout) #ordinal, 1-5 levels
unique(d.mat$Dalc) #ordinal, 1-5 levels, workday alcohol consumption
unique(d.mat$Walc) #ordinal, 1-5 levels, weekend alcohol consumption
unique(d.mat$health) #ordinal, 1-5 levels
unique(d.mat$absences) #ratio, 0- 93
hist(d.mat$absences)

```
Dependent var for Math
```{r}
hist(d.mat$G1, freq = FALSE, breaks = 10)
hist(d.mat$G2, freq = FALSE, breaks = 10)
hist(d.mat$G3, freq = FALSE, breaks = 10)


d.mat$G4 <- (0.3*d.mat$G1 + 0.3*d.mat$G2 + 0.4*d.mat$G3)/20 #computed a composite score as main dependendent variable


d.mat[1:6,-(1:30)]

hist(d.mat$G4, freq = FALSE, breaks = 12)
```

Dependent var for Port
```{r}
hist(d.port$G1, freq = FALSE, breaks = 10)
hist(d.port$G2, freq = FALSE, breaks = 10)
hist(d.port$G3, freq = FALSE, breaks = 10)


d.port$G4 <- (0.3*d.port$G1 + 0.3*d.port$G2 + 0.4*d.port$G3)/20 #computed a composite score as main dependendent variable


d.port[1:6,-(1:30)]

hist(d.port$G4, freq = FALSE, breaks = 12)
```


Studytime and results
```{r}
boxplot(G4 ~ studytime, data = d.mat, main= "Studytime and Mark for maths")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ studytime, data = d.port, main= "Studytime and Mark for Portugese")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")
```

Family supp and results
```{r}
boxplot(G4 ~ famsup, data = d.mat, main= "Studytime and Mark for maths")
boxplot(G4 ~ famsup, data = d.port, main= "Studytime and Mark for Portugese")
```
Mothers education and results
```{r}
boxplot(G4 ~ Medu, data = d.mat, main= "Mothers education and Mark for maths")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ Medu, data = d.port, main= "Mothers education and Mark for Portugese")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")
```

Fathers education and results
```{r}
boxplot(G4 ~ Fedu, data = d.mat, main= "Fathers education and Mark for maths")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ Fedu, data = d.port, main= "Fathers education and Mark for Portugese")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")
```
Mothers Job and results
```{r}
boxplot(G4 ~ Mjob, data = d.mat, main= "Mothers Job and Mark for maths")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ Mjob, data = d.port, main= "Mothers Job and Mark for Portugese")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")
```
Fathers Job and results
```{r}
boxplot(G4 ~ Fjob, data = d.mat, main= "Fathers Job and Mark for maths")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ Mjob, data = d.port, main= "Mothers Job and Mark for Portugese")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")
```

Student sex and results
```{r}
boxplot(G4 ~ sex, data = d.mat, main= "Student sex and results in Math")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ sex, data = d.port, main= "Student sex and results in Language")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")
```


Weekend alcohol consumption and results
```{r}
boxplot(G4 ~ Walc , data = d.mat, main= "Weekend alcohol consumption and results in Math")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ Walc , data = d.port, main= "Weekend alcohol consumption and results in Language")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")

plyr::count(d.mat, "Walc")
```
Weekday alcohol consumption and results
```{r}
boxplot(G4 ~ Dalc , data = d.mat, main= "Weekday alcohol consumption and results in Math")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ Dalc , data = d.port, main= "Weekday alcohol consumption and results in Language")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")

plyr::count(d.mat, "Dalc")
```

Paid for extra classes
```{r}
boxplot(G4 ~ paid , data = d.mat, main= "Paid for extra classes and results in Math")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ paid , data = d.port, main= "Paid for extra classes and results in Language")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")

plyr::count(d.mat, "paid")
```

freetime 
```{r}
boxplot(G4 ~ freetime  , data = d.mat, main= "freetime and results in Math")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ freetime  , data = d.port, main= "freetime and results in Language")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")

plyr::count(d.mat, "freetime")
```
goout  
```{r}
par(mfrow=c(1,2))
boxplot(G4 ~ goout   , data = d.mat, main= "Going out and results in Math")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ goout   , data = d.port, main= "Going out  and results in Language")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")

plyr::count(d.mat, "goout")
```



Family Status and results
```{r}
boxplot(G4 ~ Pstatus, data = d.mat, main= "Family Status and results for maths")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ Pstatus, data = d.port, main= "Family Status and results for Portugese")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")

plyr::count(d.mat$Pstatus=="A")
```

Urban/rural and results
```{r}
boxplot(G4 ~ address, data = d.mat, main= "Urban/rural and results for maths")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ address, data = d.port, main= "Urban/rural and results for Portugese")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")
```
nursery and results
```{r}
par(mfrow=c(1,2))
boxplot(G4 ~ nursery, data = d.mat, main= "nursery and results for maths")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ nursery, data = d.port, main= "nursery and results for Portugese")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")
```


School and results
```{r}
par(mfrow=c(1,2))
boxplot(G4 ~ school, data = d.mat, main= "School and results for maths")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ school, data = d.port, main= "School and results for Portugese")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")

plyr::count(d.mat$school=="GP")
```


schoolsup and results
```{r}
par(mfrow=c(1,2))
boxplot(G4 ~ schoolsup, data = d.mat, main= "schoolsup and results for maths")
abline(h=mean(d.mat$G4), col= "blue", lty="dashed")
boxplot(G4 ~ schoolsup, data = d.port, main= "schoolsup and results for Portugese")
abline(h=mean(d.port$G4), col= "blue", lty="dashed")

plyr::count(d.mat$schoolsup=="no")
```

```{r}
plot(studytime ~ traveltime, data = d.mat)  
```



```{r}
mat.pca = prcomp(~ Medu + Fedu + studytime + traveltime + age + G4, data= d.mat, retx = TRUE, center = TRUE, scale. = TRUE)
mat.pca

summary(mat.pca)

biplot(mat.pca)

```


```{r}

lm1 <- lm(G4 ~ age +address +paid + famsup + internet + Pstatus + studytime + traveltime, data = d.mat)
summary(lm1)
```
```{r}

lm2 <- lm(G4 ~ age  +paid + famsup + studytime + traveltime + romantic + famsize , data = d.mat)
summary(lm2)
```


```{r}

lm3 <- lm(G4 ~ age +log(if(absences==0,0.1)) + paid + famsup + studytime + traveltime + romantic + famsize , data = d.mat)
summary(lm3)
```



#TODO: Logistic regression  
greater than 0.5 = pass

```{r}
d.mat$G5 <- ifelse(d.mat$G4 >= 0.5, 1, 0) #1 for pass

summary(d.mat$G5)

d.mat$G6 <- ifelse(d.mat$G4 >= mean(d.mat$G4), 1, 0)

logit.model <- glm(G6 ~ studytime + traveltime + Medu + Fedu  , data= d.mat)
summary(logit.model)

```
```{r}
d.port$G5 <- ifelse(d.port$G4 >= 0.5, 1, 0) #1 for pass

summary(d.port$G5)

d.port$G6 <- ifelse(d.port$G4 >= mean(d.port$G4), 1, 0)

logit.model <- glm(G6 ~ studytime + traveltime + Medu + Fedu , data= d.port)
summary(logit.model)

```


```{r}


lm4 <- lm(G4 ~ studytime + traveltime + Medu^2 + Fedu + failures^2 + age, data= d.port)
summary(lm4)
plot(lm4)

```


#TODO: 